Subset coercions in \Coq

### 2007-01-31 papers/article/main.tex
Latest fixes

This idea is already present in mathematics. When you have
an element of subset $\{ x `: S `| P \}$, you can freely forget about the property $P$
and use any operation which is defined on $S$. Conversely, when you
want to use an operation defined on a subset, say $f : \{ x : `N `| P \} "->"
X$, you usually prove \emph{first} that you apply it only to elements 
having the desired property. For example $(f~2)$ is a correct application
only if $(P[2/x])$ is provable.
In the context of formal development of programs, such a workflow is not
entirely satisfactory because it forces one to create objects and prove
properties about them \emph{at the same time}. In \Coq{}, we are forced
to apply $f$ to an object $(\elt{\nat}{\lambda x : \nat. P}{2}{p})$ where $p$ is a proof of
$(P[2/x])$ (Figure \ref{fig:cci-subset} presents the definition of the subset type in \Coq).
We would like to be able to prove that our usage of partial functions 
is correct only after the program is written. Subsets are particularly well-suited in this respect
because they separate the objects we want to manipulate and their associated
properties. Similar treatments of subsets as a construct to separate
informative and propositional parts of an object include the work by
Nordstr\"om et al. \cite{progmartinlof} and subset types in NuPRL.

@@

Practically, this means we can use almost all existing definitions in the
\Coq environment. This map is involutive and
the identity on \Russell terms as they do not contain the constructor
$\id{elt}$ or subset projections. Moreover, this erasure function will be
defined on all \Coq terms elaborated from \Russell.
Indeed, the second projection of a
subset element will only appear in the second component of a subset
element. Otherwise, we would have used the projection directly in the
\Russell{} term, where it does not exist.
Indeed, in \PVS{} you cannot manipulate proof terms, so you have no way to create a
term of type $0 \neq 0$ from the judgement $0 : \mysubset{x}{`N}{x \neq
  0}$, nor can you actually derive the latter, as it will not type-check.
If we allowed the second projection in \Russell, that would be doable
and we would have an inconsistency.
Hence we left only the subset type forming rule (figure
\vref{fig:typing-decl-rules}) in \Russell, while the introduction and
first projection of subset elements are internalized by the coercion judgement.

@@

\subsection{Explicit coercions}
The derivation of explicit coercions (figure
\vref{fig:coerce-impl-rules-short}) is based on the
algorithmic equivalence derivations, hence we can derive a function from
the judgement which builds a coercion given two types (denote by $\coerce{`G}{T}{U}$). 
The side-conditions of the rules have not changed, so we omit them for
better readability. The judgement $\subimpl{`G}{c}{T}{U}{s}$ builds a
coercion $c$ from $T$ to $U$ in context $`G$, that is an object which
coerces an object of type $\ip{T}{`G}$ to an object of type $\ip{U}{`G}$
in context $\ipG{`G}$. We use the \lng{} types to drive the
derivation but the resulting object will be a well-typed \Coq{} term
(lemma \ref{lemma:coercion-correct}). 
% We
% could also directly use the interpreted types but that complicates the
% proof.

@@

\subsection{A little more expressiveness}
One may wonder how can we interpret \Russell terms into a system with only
$\beta\pi$-equivalence and still be a simple extension of \CIC. 
Indeed, consider the following example. Suppose we have two types $A, B$
with coercions: $c : A \suba B$ and $d : B \suba A$. Additionally, let
$P : A "->" Prop$ with introduction terms $p : \Pi x : A, P~x$ and $q :
\Pi x : B, P~x$. Consider the following well-typed \Russell term in
context $`G = x : A$: $p~x =_{P~x} q~x$. By interpretation into \CIC, we
have $\ip{P}{`G} = P$, $\ip{p}{`G} = p$ and $\ip{q}{`G} = \lambda x :
B. q~d[x] : \Pi x : B, P~d[x]$. Now the interpretation of the equality
gives:
$p~x =_{P~x} \ip{q~x}{`G} = (\lambda x : B. q~d[x]) c[x] =
q~d[c[x]]$. The right hand side has type $P~d[c[x]]$ which must be
convertible with $P~x$ for the whole term to be well-typed. This
indicates that coercions must be unique up to eta conversion and eta
conversion must be included in the definitional equality. For
example if $A = \Pi x : X.Y$ and $B = \Pi x : X'.Y'$ with $X = X'$ and
$Y = Y'$, we have $d `o c = \lambda x : X. \ctxdot~x `o \lambda x
: X'. \ctxdot~x = \lambda x : X. \ctxdot~x$ which must be definitionally
equal to $\ctxdot$.
Similarly, we must add $\eta$ rules for all introduction terms of our
language to ensure uniqueness of coercions. It is an important
property of our system, as it corresponds in practice to the determinism
of proof-obligations
generation. The complete equational theory of \CICq{} is given in figure
\vref{fig:eqccq}. It includes $\eta$, surjective pairing for dependent
sums and subsets ($\rho$) and proof-irrelevance for the second component
of subset objects ($\tau$).
Benjamin Werner has studied the addition of proof-irrelevance (rule
$\sigma$) in the Calculus of Constructions
\cite{Werner:ProofIrrelevance}. It is not a trivial extension and it has
far-reaching consequences on the model of the calculus
\cite{DBLP:conf/types/MiquelW02}. We direct the reader to these papers for further
information on the subject.
\begin{figure}[ht]
  \vspace{-1em}
  \ccqeqarr
  \vspace{-1em}
  \caption{Equational theory of \CICq}
  \label{fig:eqccq}
  \vspace{-1em}
\end{figure}

@@

\paragraph{Acknowledgements}
I would like to thank Christine Paulin-Mohring for directing this work and
Jean-Christophe Filli\^atre for discussions on previous versions of this paper.
\bibliography{my,bib-joehurd,barras,pvs-bib,bcp,Luo,biblio-subset,%
cparent,demons,demons2,demons3}
\bibliographystyle{splncs}
% \renewcommand{\thefootnote}{}
% \footnotetext{This article was typeset in \LaTeX~using the
%   \texttt{Concrete Math} font}

>>>>

This idea is already present in mathematics. When you have
an element of subset $\{ x `: S `| P \}$, you can freely forget about the property $P$
and use any operation which is defined on $S$. Conversely, when you
want to use an operation defined on a subset, say $f : \{ x : `N `| P \} "->"
X$, you usually prove \emph{first} that you apply it only to elements 
having the desired property. For example $(f~2)$ is a correct application
only if $(P[2/x])$ is provable.
In the context of formal development of programs, such a workflow is not
entirely satisfactory because it forces one to create objects and prove
properties about them \emph{at the same time}. In \Coq{}, we are forced
to apply $f$ to an object $(\elt{\nat}{\lambda x : \nat. P}{2}{p})$ where $p$ is a proof of
$(P[2/x])$ (Figure \ref{fig:cci-subset} presents the definition of the subset type in \Coq).
We would like to be able to prove that our usage of partial functions 
is correct only after the program is written. Subsets are particularly well-suited in this respect
because they separate the objects we want to manipulate and their associated
properties. Similar treatments of subsets as a construct to separate
informative and propositional parts of an object include the work by
Nordstr\"om et al. \cite{progmartinlof} and subset types in NuPRL \cite{constable86nuprl}.

@@

Practically, this means we can use almost all existing definitions in the
\Coq environment. This map is involutive and
the identity on \Russell terms as they do not contain the constructor
$\id{elt}$ or subset projections. Moreover, this erasure function will be
defined on all \Coq terms elaborated from \Russell.
Indeed, the second projection of a
subset element will only appear in the second component of a subset
element. Otherwise, we would have used the projection directly in the
\Russell{} term, where it does not exist.
Indeed, in \PVS{} you cannot manipulate proof terms, so you have no way to create a
term of type $0 \neq 0$ from the judgement $0 : \mysubset{x}{`N}{x \neq
  0}$. If we allowed the second projection in \Russell, that would be doable
and we would have an inconsistency.
Hence we left only the subset type forming rule (figure
\vref{fig:typing-decl-rules}) in \Russell, while the introduction and
first projection of subset elements are internalized by the coercion judgement.

@@

\subsection{Explicit coercions}
The derivation of explicit coercions (figure
\vref{fig:coerce-impl-rules-short}) is based on the
algorithmic equivalence derivations.  
The side-conditions of the rules have not changed, so we omit them for
better readability. The judgement $\subimpl{`G}{c}{T}{U}{s}$ builds a
coercion $c$ from $\ip{T}{`G}$ to $\ip{U}{`G}$ in context $\ipG{`G}$.
We use the \lng{} types to drive the
derivation but the resulting object $c$ will be a well-typed \Coq{} term
(lemma \ref{lemma:coercion-correct}). We denote $\coerce{`G}{T}{U}$
the function which builds $c$ given $`G$, $T$ and $U$.
% We
% could also directly use the interpreted types but that complicates the
% proof.

@@

\subsection{A little more expressiveness}
One may wonder how can we interpret \Russell terms into a system with only
$\beta\pi$-equivalence and still be a simple extension of \CIC. 
Indeed, consider the following example. Suppose we have two types $A, B$
with coercions: $c : A \suba B$ and $d : B \suba A$. Additionally, let
$P : A "->" Prop$ with introduction terms $p : \Pi x : A, P~x$ and $q :
\Pi x : B, P~x$. Consider the following well-typed \Russell term in
context $`G = x : A$: $p~x =_{P~x} q~x$. By interpretation into \CIC, we
have $\ip{P}{`G} = P$, $\ip{p}{`G} = p$ and $\ip{q}{`G} = \lambda x :
B. q~d[x] : \Pi x : B, P~d[x]$. Now the interpretation of the equality
gives:
$p~x =_{P~x} \ip{q~x}{`G}$ where $\ip{q~x}{`G} `= (\lambda x : B. q~d[x]) c[x] `=
q~d[c[x]]$. The right hand side has type $P~d[c[x]]$ which must be
convertible with $P~x$ for the whole term to be well-typed. This
indicates that coercions must be unique (so that $d `o c `= id_A `=
\ctxdot$), which implies that we have to include $\eta$ 
conversion in the definitional equality. For
example if $A = \Pi x : X.Y$ and $B = \Pi x : X'.Y'$ with $X = X'$ and
$Y = Y'$, we have $d `o c = \lambda x : X. \ctxdot~x `o \lambda x
: X'. \ctxdot~x = \lambda x : X. \ctxdot~x$ which must be definitionally
equal to $\ctxdot$.
Similarly, we must add $\eta$ rules for all introduction terms of our
language to ensure uniqueness of coercions. It is an important
property of our system, as it corresponds in practice to the determinism
of proof-obligations
generation. The complete equational theory of \CICq{} is given in figure
\vref{fig:eqccq}. It includes $\eta$, surjective pairing for dependent
sums and subsets ($\rho$) and proof-irrelevance for the second component
of subset objects ($\tau$) and is sufficient to ensure uniqueness of coercions.
Benjamin Werner has studied the addition of proof-irrelevance (rule
$\sigma$) in the Calculus of Constructions
\cite{Werner:ProofIrrelevance}. It is not a trivial extension and it has
far-reaching consequences on the model of the calculus
\cite{DBLP:conf/types/MiquelW02}. We direct the reader to these papers for further
information on the subject.
\begin{figure}[ht]
  \vspace{-1em}
  \ccqeqarr
  \vspace{-1em}
  \caption{Equational theory of \CICq}
  \label{fig:eqccq}
  \vspace{-1em}
\end{figure}

@@

\paragraph{Acknowledgements}
We would like to thank Christine Paulin-Mohring for directing this work and
Jean-Christophe Filli\^atre for discussions on previous versions of this
paper. We thank the anonymous referees for their insightful comments.
\bibliography{my,bib-joehurd,barras,pvs-bib,bcp,Luo,biblio-subset,%
cparent,demons,demons2,demons3}
\bibliographystyle{splncs}
% \renewcommand{\thefootnote}{}
% \footnotetext{This article was typeset in \LaTeX~using the
%   \texttt{Concrete Math} font}
